skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-07-15 paper/paper.Rmd
typos, funding statement

Inferring familial relationships between individuals using genetic data is a common practice in population genetics, medical genetics, and forensics. There are multiple approaches to estimating relatedness between samples, including genome-wide measures such as those implemented in Plink [@purcell2007] or KING [@manichaikul2010], and methods that rely on identity by descent (IBD) segment detection such as GERMLINE [@gusev2009], hap-IBD [@zhou2020], and IBIS [@seidman2020]. Recent efforts focusing on benchmarking these methods [@ramstetter2017; @vries2021] have been aided by tools for simulating pedigrees and genome-wide SNP data [@caballero2019]. Analyzing results from genome-wide SNP-based kinship analysis or comparing analyses to simulated data for benchmarking have to this point required writing one-off analysis functions or utility scripts that are seldom distributed with robust documentation, test suites, or narrative examples of usage. There is a need in the field for a well-documented software package with a consistent design and API that contains functions to assist with downstream manipulation, benchmarking, and analysis of SNP-based kinship assessment methods. Here we present the skater package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation with **R**.

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the function `read_fam()` to read in a PLINK-formatted .fam file, and another function `fam2ped()` to convert the content into a pedigree object as a nested tibble with one row per family. All pedigree processing from skater internally leverages a data structure from the kinship2 package [@kinship2]. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to known degrees of relatedness. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette (see Supplementary Material), and may be a preferrable measure of classification accuracy when benchmarking relationship degree estimation. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second degree) than misclassifications at more distant relationships (e.g., classifying a fourth-degree relative pair as fifth-degree).

@@

Tools such as hap-IBD [@zhou2020], and IBIS [@seidman2020] detect shared IBD segments between individuals. The skater package includes functionality to take those IBD segments, compute shared genomic centimorgan (cM) length, and convert that shared cM to a kinship coefficient. In addition to inferred segments, these functions can estimate "truth" kinship from simulated IBD segments [@caballero2019]. The `read_ibd()` function reads pairwise IBD segments from IBD inference tools and from simulated IBD segments. The `read_map()` function reads in genetic map in a standard format which is required to translate the total centimorgans shared IBD to a kinship coefficient using the `ibd2kin()` function.

@@

**UNT FOLKS: NEED YOUR INPUT HERE!** The following rules should be followed:

-   The sentence should begin: 'This work was supported by ...' -
-   The full official funding agency name should be given, i.e. 'National Institutes of Health', not 'NIH' (full RIN-approved list of UK funding agencies)
-   Grant numbers should be given in brackets as follows: '[grant number xxxx]'
-   Multiple grant numbers should be separated by a comma as follows: '[grant numbers xxxx, yyyy]'
-   Agencies should be separated by a semi-colon (plus 'and' before the last funding agency)
-   Where individuals need to be specified for certain sources of funding the following text should be added after the relevant agency or grant number 'to [author initials]'.

An example is given here: 'This work was supported by the National Institutes of Health [AA123456 to C.S., BB765432 to M.H.]; and the Alcohol & Education Research Council [hfygr667789].'

Oxford Journals will deposit all NIH-funded articles in PubMed Central. See Depositing articles in repositories -- information for authors for details. Authors must ensure that manuscripts are clearly indicated as NIH-funded using the guidelines above.

>>>>

Inferring familial relationships between individuals using genetic data is a common practice in population genetics, medical genetics, and forensics. There are multiple approaches to estimating relatedness between samples, including genome-wide measures, such as those implemented in Plink [@purcell2007] or KING [@manichaikul2010], and methods that rely on identity by descent (IBD) segment detection, such as GERMLINE [@gusev2009], hap-IBD [@zhou2020], and IBIS [@seidman2020]. Recent efforts focusing on benchmarking these methods [@ramstetter2017; @vries2021] have been aided by tools for simulating pedigrees and genome-wide SNP data [@caballero2019]. Analyzing results from genome-wide SNP-based kinship analysis or comparing analyses to simulated data for benchmarking have to this point required writing one-off analysis functions or utility scripts that are seldom distributed with robust documentation, test suites, or narrative examples of usage. There is a need in the field for a well-documented software package with a consistent design and API that contains functions to assist with downstream manipulation, benchmarking, and analysis of SNP-based kinship assessment methods. Here we present the skater package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation with **R**.

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the function `read_fam()` to read in a PLINK-formatted .fam file and another function `fam2ped()` to convert the content into a pedigree object as a nested tibble with one row per family. All pedigree processing from skater internally leverages a data structure from the kinship2 package [@kinship2]. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to known degrees of relatedness. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette (see Supplementary Material) and may be a preferable measure of classification accuracy when benchmarking relationship degree estimation. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second-degree) than misclassifications at more distant relationships (e.g., classifying a fourth-degree relative pair as fifth-degree).

@@

Tools such as hap-IBD [@zhou2020], and IBIS [@seidman2020] detect shared IBD segments between individuals. The skater package includes functionality to take those IBD segments, compute shared genomic centimorgan (cM) length, and converts that shared cM to a kinship coefficient. In addition to inferred segments, these functions can estimate "truth" kinship from simulated IBD segments [@caballero2019]. The `read_ibd()` function reads pairwise IBD segments from IBD inference tools and from simulated IBD segments. The `read_map()` function reads in genetic map in a standard format which is required to translate the total centimorgans shared IBD to a kinship coefficient using the `ibd2kin()` function.

@@

This work was supported in part by award 2019-DU-BX-0046 (Dense DNA Data for Enhanced Missing Persons Identification) to B.B., awarded by the National Institute of Justice, Office of Justice Programs, U.S. Department of Justice and by internal funds from the Center for Human Identification. The opinions, findings, and conclusions
or recommendations expressed are those of the authors and do not necessarily reflect those of the U.S. Department of Justice.

<!-- -   The sentence should begin: 'This work was supported by ...' - -->
<!-- -   The full official funding agency name should be given, i.e. 'National Institutes of Health', not 'NIH' (full RIN-approved list of UK funding agencies) -->
<!-- -   Grant numbers should be given in brackets as follows: '[grant number xxxx]' -->
<!-- -   Multiple grant numbers should be separated by a comma as follows: '[grant numbers xxxx, yyyy]' -->
<!-- -   Agencies should be separated by a semi-colon (plus 'and' before the last funding agency) -->
<!-- -   Where individuals need to be specified for certain sources of funding the following text should be added after the relevant agency or grant number 'to [author initials]'. -->

<!-- An example is given here: 'This work was supported by the National Institutes of Health [AA123456 to C.S., BB765432 to M.H.]; and the Alcohol & Education Research Council [hfygr667789].' -->

<!-- Oxford Journals will deposit all NIH-funded articles in PubMed Central. See Depositing articles in repositories -- information for authors for details. Authors must ensure that manuscripts are clearly indicated as NIH-funded using the guidelines above. -->
